# Summary {findings-summary}

```{r}
#| label: setup

library(tidyverse)
library(kableExtra)
library(hdatools)
library(formattable)
library(sf)
library(ggtext)
library(scales)
library(ggspatial)
library(geojsonsf)

```

Out of the over 4 million parcels in Virginia, **22,453** were identified as likely owned by a faith-based organization --- with a 95% confidence level that the true accuracy of the entire dataset lies between 99.03% and 100%.

```{r}
#| label: fig-map
#| fig-cap: "Parcels in Virginia owned by faith-based organizations"
#| fig-asp: 0.55

faith_mapped <- read_rds("data/faith_mapped.rds") |> 
  distinct(ll_uuid, .keep_all = TRUE) |> 
  st_centroid()

va_cty <- geojson_sf("data/geo/pdc_va.geojson") 

low_cov <- c("Alleghany", "Cumberland", "Dinwiddie", "Goochland", "Rockbridge", "Washington")

ggplot() +
  geom_sf(
    data = va_cty,
    fill = "grey90",
    color = "grey95" 
  ) +
  geom_sf(
    data = filter(va_cty, Name %in% low_cov),
    fill = "grey90",
    color = "grey50",
    linewidth = 0.5,
    linetype = 5
  ) +
  geom_sf(
    data = faith_mapped,
    size = 0.5,
    color = "#259591",
    alpha = 0.1
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    title = "Parcels in Virginia owned by faith-based organizations",
    subtitle = "Limited data coverage in outlined counties"
  ) +
  theme_hfv(
    panel.grid = element_blank(),
    axis.text = element_blank()
  )

```


These findings were based on a keyword search with a process of retention and elimination based on known faith-based and secular entities. Some faith-based properties could be absent from the analysis due to terms or phrases not accounted for.

However, the results likely represent a significant, if not an overwhelming majority, of parcels owned by faith-based organizations in the Commonwealth of Virginia. In total, these parcels amount to 74,116 acres of land.

::: {.callout-tip}
### How large is 74,116 acres?
To put this into context, the City of Richmond is roughly 40,000 acres, meaning the identified parcels amount to nearly *twice the size* of Richmond.
:::

A histogram of the lot sizes of the 22,453 parcels shows that an overwhelming majority of these parcels are more than 0.1 acres (approximately 4,350 square feet). 

```{r}
#| label: fig-lots
#| fig-cap: "Distribution of lot sizes"

faith_parcels <- faith_mapped |> 
  st_drop_geometry()

lot_size_label <- function(x) {
  ifelse(
    x < 0.1,
    paste(format(round(x * 43560), big.mark = ","), "sqft"),
    paste(formatC(x, digits = 3), "acres")
  )
}

median_lot <- median(faith_parcels$acres)

median_label <- lot_size_label(median_lot)

lb <- 10^(seq(log10(0.01), log10(100), length.out = 31))

lb_plot <- c(
  500/43560, 1000/43560, 2000/43560,
  0.1, 0.25, 0.5, 1, 2, 5, 10, 25, 50, 100
  )

ggplot(faith_parcels, aes(y = acres)) +
  geom_histogram(
    breaks = lb,
    fill = "#011E41",
    color = NA,
    alpha = 0.2
  ) +
  geom_density(
    aes(x = ..count.. * 0.14),
    color = "#011E41", 
    size = 0.5,
    alpha = 0.7
  ) +
  # Add vertical line for median
  geom_hline(
    yintercept = median_lot,
    color = "#8B85CA",
    linetype = "dashed",
    size = 0.75
  ) +
  annotate(
    "text",
    y = median_lot,
    x = 1850,
    label = paste("Median:", median_label),
    color = "#8B85CA",
    size = 3,
    hjust = 1,
    vjust = -1,
    fontface = "bold"
  ) +
  coord_cartesian(ylim = c(100, 0.01)) +
  scale_x_continuous(
    limits = c(0, 1900),
    expand = c(0.01, 0.01),
    labels = label_comma()
  ) +
  scale_y_continuous(
    trans = c("log10", "reverse"),
    breaks = lb_plot,
    expand = c(0.02, 0.02),
    labels = lot_size_label
  ) +
  labs(
    title = "Distribution of lot sizes",
    subtitle = "Small lots shown in square feet, larger lots in acres",
    x = "Number of parcels"
  ) +
  theme_hfv(
    flip_gridlines = T,
    axis.title.x = element_text(hjust = 0, margin = margin(t = 6)),
    axis.text.y = element_text(vjust = 0.5)
  )


```

SB 822 or any similar legislation would encourage multifamily development whose lot size requirements depend on a variety of factors. The bill introduced in 2024 allowed for a density of 40 units per acre and a height of one story *or* 15 feet above the maximum allowable height for residential use according to the locality's current zoning.

The median lot size is 23,509 square feet, which is approximately 0.54 acres. In many circumstances, this lot size is only viable for a small multifamily building of fewer than 20 units, particularly in urban areas served by public water and sewer.

However, there are still significant shares of these parcels that could theoretically support larger residential development projects. About 38% are larger than 1 acre, and 25% are larger than 2 acres.

```{r}
#| label: tbl-lot-sum
#| tbl-cap: "Share of parcels by lot size"

lot_sum <- faith_parcels |> 
  summarise(
    `More than 0.25 acres` = mean(acres > 0.25),
    `More than 0.5 acres` = mean(acres > 0.5),
    `More than 1 acre` = mean(acres > 1),
    `More than 2 acres` = mean(acres > 2),
    `More than 5 acres` = mean(acres > 5)
  )

lot_sum |> 
  pivot_longer(
    cols = everything(),
    names_to = "Lot size",
    values_to = "Percent"
  ) |> 
  mutate(Percent = percent(Percent, digits = 1)) |> 
  kbl(
    align = "lr"
  ) |> 
  kable_styling(
    full_width = F,
    bootstrap_options = c("striped")
  )

```

