# Findings {part-2}

```{r}
#| label: setup

library(tidyverse)
library(kableExtra)
library(hdatools)
library(sf)
library(geojsonsf)
library(ggtext)
library(scales)
library(leaflet)
library(htmlwidgets)
library(classInt)
library(htmltools)

```

## Summary

Out of the over 4 million parcels in Virginia, 22,453 were identified as likely owned by a faith-based organization --- with a 95% confidence level that the true accuracy of the entire dataset lies between 99.03% and 100%. These findings were based on a keyword search with a process of retention and elimination based on known faith-based and secular entities. Some faith-based properties could be absent from the analysis due to terms or phrases not accounted for. However, the following represents a significant, if not an overwhelming majority, of parcels owned by faith-based organizations in the Commonwealth of Virginia. In total, these parcels amount to 74,116 acres of land. To put this into context, the City of Richmond is roughly 40,000 acres, meaning the identified parcels amount to nearly twice the size of Richmond.

### By Planning District Commission

To investigate geographic patterns further, HFV utilized Virginia's 21 planning district commissions (PDCs) to assign parcels to a geographic region. This is helpful because not all jurisdictions in Virginia are part of a metropolitan or micropolitan statistical area, whereas all jurisdictions are represented by at least one of the 21 PDCs. In addition, PDC boundaries represent an area wherein there are distinct socioeconomic patterns.

```{r}
#| label: fig-pdc-map
#| fig-cap: "Map of Virginia PDCs"

# Insert map of Virginia Planning District Commissions

pdc <- geojson_sf("data/geo/pdc_va.geojson") 


# Create a custom color palette with 21 distinct colors
colors_21 <- c(
  "#FF0000", "#00FF00", "#0000FF", "#FFFF00", "#FF00FF", "#00FFFF", "#800000",
  "#008000", "#000080", "#808000", "#800080", "#008080", "#FFA500", "#FFC0CB",
  "#A52A2A", "#DEB887", "#5F9EA0", "#7FFF00", "#D2691E", "#FF7F50", "#6495ED"
)

# Create a color palette function
pal <- colorFactor(
  palette = colors_21,
  domain = pdc$PDC_Primary_Name
)

# Create a custom tooltip
tooltip_content <- paste0(
  "<strong>Jurisdiction:</strong> ", pdc$Name, "<br/>",
  "<strong>PDC:</strong> ", pdc$PDC_Primary_Name
) %>% lapply(HTML)

# Create the leaflet map
map <- leaflet(pdc) %>%
  addProviderTiles("CartoDB.Positron") |> 
  addPolygons(
    fillColor = ~pal(PDC_Primary_Name),
    weight = 1,
    opacity = 0.6,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 1,
      color = "#666",
      dashArray = "",
      fillOpacity = 1,
      bringToFront = TRUE),
    label = tooltip_content,
    labelOptions = labelOptions(
      style = list(
        "font-family" = "sans-serif",
        "font-size" = "12px",
        padding = "10px"
      ),
      textsize = "15px",
      direction = "auto",
      opacity = 0.9,
      offset = c(0, -10),
      sticky = TRUE  # Make tooltip follow the mouse
    ))

# Display the map
map


```

The Hampton Roads region exceeds its peers in the number of parcels owned by faith-based organizations at 3,332. The LENOWISCO and PlanRVA regions follow at just over 1,000 fewer parcels - 2,050 and 2,006, respectively.

```{r}
#| label: fig-pdc
#| fig-cap: "Total faith-based owned parcels by PDC"

faith_mapped <- read_rds("data/faith_mapped.rds") |> 
  st_drop_geometry() |> 
  distinct(ll_uuid, .keep_all = TRUE) |> 
  mutate(pdc = case_when(
    jurisdiction == "Franklin County" ~ "Roanoke Valley-Alleghany",
    TRUE ~ pdc
  ))


faith_summary_pdc <- faith_mapped |> 
  group_by(pdc) |> 
  summarise(total_acres = as.numeric(format(sum(acres), scientific = FALSE)),
            avg_parcel_size = mean(acres),
            count = as.numeric(n_distinct(ll_uuid)))

ggplot(faith_summary_pdc,
       aes(x = reorder(pdc, count),
           y = count)) +
  geom_col(fill = "#011E41") +
  geom_text(
    aes(label = scales::comma(round(count, 0))), 
    color = "#011E41", 
    hjust = -0.2
    ) +
  coord_flip() +
  theme_hfv() +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Total faith-based owned parcels",
       caption = "**Source:** HousingForward Virginia & LOVELAND Technologies, Regrid 2024 Dataset.")


```

These rankings change when we look specifically at acreage. The Northern Shenandoah Valley region rises to the top of list at just over 14,000 acres of land owned by faith-based organizations, while the PlanRVA region follows at 10,246. Hampton Roads still stays in the top three, but comes in at 7,602 acres. The LENOWISCO region falls down the list to the 17th region with the most acreage at only 1,606 acres. While the LENOWISCO region contains significantly more parcels than several of its peers, those parcels are undoubtedly smaller in area.

```{r}
#| label: fig-pdc-acres
#| fig-cap: "Total acres of faith-based owned parcels by PDC"

faith_mapped <- read_rds("data/faith_mapped.rds") |> 
  st_drop_geometry() |> 
  distinct(ll_uuid, .keep_all = TRUE) |> 
  mutate(pdc = case_when(
    jurisdiction == "Franklin County" ~ "Roanoke Valley-Alleghany",
    TRUE ~ pdc
  ))

faith_summary_pdc <- faith_mapped |> 
  group_by(pdc) |> 
  summarise(total_acres = as.numeric(format(sum(acres), scientific = FALSE)),
            avg_parcel_size = mean(acres),
            count = as.numeric(n_distinct(ll_uuid)))

# sum(faith_summary_pdc$total_acres)

ggplot(faith_summary_pdc,
       aes(x = reorder(pdc, total_acres),
           y = total_acres)) +
  geom_col(fill = "#011E41") +
  geom_text(
    aes(label = scales::comma(round(total_acres, 0))), 
    color = "#011E41", 
    hjust = -0.2
    ) +
  coord_flip() +
  theme_hfv() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Total acres of faith organization owned land",
       caption = "**Source:** HousingForward Virginia & LOVELAND Technologies, Regrid 2024 Dataset.")

```

Exploring individual parcel size gives a better understanding of opportunities for viable residential development. To do this, we explore the median parcel size rather than the average in order to account for outliers that may skew the results. By looking at the median parcel size, the rankings change once again. The Thomas Jefferson region has the largest parcels with a median size of 1.74 acres - just under the size of a FIFA standard soccer field (1.76 acres). The Northern Neck follows at 1.55 acres and the next six regions sit between 1.3 and 0.92 acres. The remaining regions have a median parcel size below 0.75 acres - with the LENOWISCO region have the smallest median parcel size at under 0.10 acres.

```{r}
#| label: fig-pdc-med-acres
#| fig-cap: "Median parcel size by PDC"

faith_mapped <- read_rds("data/faith_mapped.rds") |> 
  st_drop_geometry() |> 
  distinct(ll_uuid, .keep_all = TRUE) |> 
  mutate(pdc = case_when(
    jurisdiction == "Franklin County" ~ "Roanoke Valley-Alleghany",
    TRUE ~ pdc
  ))

faith_summary_pdc <- faith_mapped |> 
  group_by(pdc) |> 
  summarise(total_acres = as.numeric(format(sum(acres), scientific = FALSE)),
            med_parcel_size = median(acres),
            count = as.numeric(n_distinct(ll_uuid)))

ggplot(faith_summary_pdc,
       aes(x = reorder(pdc, med_parcel_size),
           y = med_parcel_size)) +
  geom_col(fill = "#40c0c0") +
    geom_text(aes(label = round(med_parcel_size, 2)), color = "#FFFFFF", hjust = 1.1) +
  coord_flip() +
  theme_hfv() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)), limits = c(0,2))  +
  labs(title = "Median parcel size of faith-based owned land by PDC",
       caption = "**Source:** HousingForward Virginia & LOVELAND Technologies, Regrid 2024 Dataset.")



```

At the local level, we can identify jurisdictions with significant acreage of faith-based owned parcels. Based on median parcel size, Goochland County leads its nearest peer by over 1.5 acres with a median parcel size of 7.18 acres. Cumberland County comes in at 5.5 acres. The PlanRVA region is well-represented among jurisdictions with large median parcel sizes - with Goochland, Powhatan, Henrico, New Kent, Hanover, and Charles City represented (6 of the 8 counties/cities within the PDC).

```{r}
#| label: fig-local-count
#| fig-cap: "Top 15: Median parcel size by jurisdiction"

faith_mapped <- read_rds("data/faith_mapped.rds")|> 
  st_drop_geometry() |> 
  distinct(ll_uuid, .keep_all = TRUE) |> 
  mutate(pdc = case_when(
    jurisdiction == "Franklin County" ~ "Roanoke Valley-Alleghany",
    TRUE ~ pdc
  ))

faith_summary_local <- faith_mapped %>% 
  group_by(jurisdiction, pdc) %>% 
  summarise(
    total_acres = sum(acres, na.rm = TRUE),
    med_parcel_size = median(acres, na.rm = TRUE),
    count = n_distinct(ll_uuid),
    .groups = "drop"
  ) %>% 
  mutate(
    jurisdiction = str_remove(jurisdiction, " County"),
    total_acres = round(total_acres, 2),
    med_parcel_size = round(med_parcel_size, 2)
  ) %>% 
  arrange(desc(med_parcel_size)) %>%
  slice_head(n = 20)

ggplot(faith_summary_local,
       aes(x = reorder(jurisdiction, med_parcel_size),
           y = med_parcel_size,
           fill = pdc)) +
  geom_col(fill = "#40c0c0") +
  geom_text(aes(label = med_parcel_size), color = "#FFFFFF", hjust = 1.1) +
  coord_flip() +
  theme_hfv()

```

The top ten organizations who own the most land in jurisdictions along the I-95/I-64 corridor are listed below.The Presbyterian League of the Presbytery of Eastern Virginia owns the most acreage with 12 parcels that amount to 3,429 acres in New Kent County. None of the top owners are located within a major city, but are largely in rural communities.

```{r}
#| label: fig-local-own
#| fig-cap: "Top 5: Parcel count by owners in regions along I-95/I-64 "

faith_mapped <- read_rds("data/faith_mapped.rds")|> 
  st_drop_geometry() |> 
  distinct(ll_uuid, .keep_all = TRUE) |> 
  filter(
    pdc == "Hampton Roads" | 
      pdc == "PlanRVA" |
      pdc == "Northern Virginia" |
      pdc == "George Washington")

faith_summary_local <- faith_mapped %>% 
  group_by(owner, jurisdiction) %>% 
  summarise(
    total_acres = sum(acres, na.rm = TRUE),
    med_parcel_size = median(acres, na.rm = TRUE),
    count = n_distinct(ll_uuid),
    .groups = "drop"
  ) %>% 
  mutate(
    jurisdiction = str_remove(jurisdiction, " County"),
    total_acres = round(total_acres, 2),
    med_parcel_size = round(med_parcel_size, 2)
  ) %>% 
  arrange(desc(total_acres)) %>%
  slice_head(n = 10) |> 
  mutate(owner = str_to_title(owner))


faith_summary_local |> 
  kbl(col.names = c("Owner", "Jurisdiction", "Total Acres", "Median Parcel Size", "Number of Parcels"), align = "lcccc") |> 
  kable_styling(bootstrap_options = c("striped", "condensed"))


```

Further exploration of ownership in specific jurisdictions of interest may yield results that would facilitate conversation about current use and future plans for parcels. The current investigation would only be enhanced with major groundtruthing efforts and outreach to individual parcel owners.

## General Assembly Districts

The introduction of the Faith in Housing bill into the Virginia General Assembly presents an opportunity to also explore faith-based parcel ownership at based on house and senate districts. This provides state legislatures with a unique perspective on aspects of faith-based parcel ownership within their districts. There are 100 House Districts and 40 Senate Districts in the Virginia General Assembly. District lines were \[recently redrawn\])https://www.vpap.org/redistricting/) in 2021 by the Supreme Court of Virginia after Virginia's Redistricting Commission never came to an agreement.

### House Districts

Based on House Districts, the majority of faith-based parcel ownership is in southern and southwest Virginia. House Districts 45 and 46 contain the largest number of parcels owned by faith organizations, 2,080 and 1,199, respectively. They are followed by House District 49 (928 parcels).

```{r}
#| label: fig-parcels-hd
#| fig-cap: "Total parcels by House District"


# hd <- read_rds("data/rds/faith_house.rds")
# 
# # Function to create color palette based on Jenks breaks
# create_jenks_palette <- function(data, n_classes = 5) {
#   # Remove any NA values
#   data <- data[!is.na(data)]
#   
#   # Calculate Jenks breaks
#   breaks <- classIntervals(data, n = n_classes, style = "jenks")
#   
#   # Create a color palette
#   pal <- colorBin(
#     palette = "YlOrRd",
#     domain = data,
#     bins = breaks$brks,
#     pretty = FALSE
#   )
#   
#   return(pal)
# }
# 
# # Create the Jenks breaks color palette
# jenks_pal <- create_jenks_palette(hd$faith_house_count)
#   
# 
# # Create the leaflet map
# map_jenks <- leaflet(hd) %>%
#   addProviderTiles("CartoDB.Positron") %>%
#   addPolygons(
#     fillColor = ~jenks_pal(faith_house_count),
#     fillOpacity = 0.7,
#     color = "grey",
#     weight = 1,
#     popup = ~paste("House District:", NAMELSAD, "<br>Parcels:", label_count)
#   ) %>%
#   addLegend(
#     pal = jenks_pal,
#     values = hd$faith_house_count,
#     title = "Total Parcels",
#     position = "topleft"
#   )
# 
# saveWidget(map_jenks, "data/geo/hd_count.html")

# Display the map

knitr::include_url("data/geo/hd_count.html", height = "500px")


```

When focused on total acreage of faith-owned parcels, House District 33 and and House District 71 exceed their peers. House District 33's large amount of faith-owned land is due greatly to the presence of Shrine Mont, a large retreat and conference center located in Orkney Springs, Virginia and owned by the Episcopal Diocese.

```{r}
#| label: fig-acres-hd
#| fig-cap: "Total acres by House District"


# hd <- read_rds("data/rds/faith_house.rds")
# 
# # Function to create color palette based on Jenks breaks
# create_jenks_palette <- function(data, n_classes = 5) {
#   # Remove any NA values
#   data <- data[!is.na(data)]
#   
#   # Calculate Jenks breaks
#   breaks <- classIntervals(data, n = n_classes, style = "jenks")
#   
#   # Create a color palette
#   pal <- colorBin(
#     palette = "YlGn",
#     domain = data,
#     bins = breaks$brks,
#     pretty = FALSE
#   )
#   
#   return(pal)
# }
# 
# # Create the Jenks breaks color palette
# jenks_pal <- create_jenks_palette(hd$faith_house_total_acres)
# 
# # Create the leaflet map
# map_jenks <- leaflet(hd) %>%
#   addProviderTiles("CartoDB.Positron") %>%
#   addPolygons(
#     fillColor = ~jenks_pal(faith_house_total_acres),
#     fillOpacity = 0.7,
#     color = "grey",
#     weight = 1,
#     popup = ~paste("House District:", NAMELSAD, "<br>Value:", label_acres)
#   ) %>%
#   addLegend(
#     pal = jenks_pal,
#     values = hd$faith_house_total_acres,
#     title = "Total Acres",
#     position = "topleft"
#   )
# 
# # Display the map
# saveWidget(map_jenks, "data/geo/hd_acres.html")

knitr::include_url("data/geo/hd_acres.html", height = "500px")

```

The median parcel size can better help us understand areas where there is adequate, contiguous land that could be developed. Based on House Districts, large parcels are located in House District 88 (9.54 acres), House District 28 (9.26 acres), and House District 26 (7.14). The presence of smaller median parcel size in a district does not necessarily mean that properties are too small to develop, but could be an indication of multiple adjacent parcels under the same ownership. However, the results provide us with a preliminary assessment of typical parcel size in each district.

```{r}
#| label: fig-size-hd
#| fig-cap: "Median parcel size by House District"


# hd <- read_rds("data/rds/faith_house.rds")
# 
# # Function to create color palette based on Jenks breaks
# create_jenks_palette <- function(data, n_classes = 5) {
#   # Remove any NA values
#   data <- data[!is.na(data)]
#   
#   # Calculate Jenks breaks
#   breaks <- classIntervals(data, n = n_classes, style = "jenks")
#   
#   # Create a color palette
#   pal <- colorBin(
#     palette = "YlGnBu",
#     domain = data,
#     bins = breaks$brks,
#     pretty = FALSE
#   )
#   
#   return(pal)
# }
# 
# # Create the Jenks breaks color palette
# jenks_pal <- create_jenks_palette(hd$med_parcel_size)
# 
# # Create the leaflet map
# map_jenks <- leaflet(hd) %>%
#   addProviderTiles("CartoDB.Positron") %>%
#   addPolygons(
#     fillColor = ~jenks_pal(med_parcel_size),
#     fillOpacity = 0.7,
#     color = "grey",
#     weight = 1,
#     popup = ~paste("House District:", NAMELSAD, "<br>Value:", label_size)
#   ) %>%
#   addLegend(
#     pal = jenks_pal,
#     values = hd$med_parcel_size,
#     title = "Median Parcel Size",
#     position = "topleft"
#   )
# 
# # Display the map
# saveWidget(map_jenks, "data/geo/hd_size.html")

knitr::include_url("data/geo/hd_size.html", height = "500px")

```

### Senate Districts

Senate Districts undoubtedly show similar patterns as House Districts. Senate Districts 40 (2,545) and 15 (1,727), in southwest and southern Virginia, contain the most faith-owned parcels. They are followed by other Senate Districts in the south and southwest part of the state.

```{r}
#| label: fig-parcels-sd
#| fig-cap: "Total parcels by Senate District"

# sd <- read_rds("data/rds/faith_senate.rds")
# 
# # Function to create color palette based on Jenks breaks
# create_jenks_palette <- function(data, n_classes = 5) {
#   # Remove any NA values
#   data <- data[!is.na(data)]
#   
#   # Calculate Jenks breaks
#   breaks <- classIntervals(data, n = n_classes, style = "jenks")
#   
#   # Create a color palette
#   pal <- colorBin(
#     palette = "YlOrRd",
#     domain = data,
#     bins = breaks$brks,
#     pretty = FALSE
#   )
#   
#   return(pal)
# }
# 
# # Create the Jenks breaks color palette
# jenks_pal <- create_jenks_palette(sd$faith_senate_count)
#   
# 
# # Create the leaflet map
# map_jenks <- leaflet(sd) %>%
#   addProviderTiles("CartoDB.Positron") %>%
#   addPolygons(
#     fillColor = ~jenks_pal(faith_senate_count),
#     fillOpacity = 0.7,
#     color = "grey",
#     weight = 1,
#     popup = ~paste("Senate District:", NAMELSAD, "<br>Parcels:", label_count)
#   ) %>%
#   addLegend(
#     pal = jenks_pal,
#     values = sd$faith_senate_count,
#     title = "Total Parcels",
#     position = "topleft"
#   )
# 
# # Display the map
# saveWidget(map_jenks, "data/geo/sd_count.html")

knitr::include_url("data/geo/sd_count.html", height = "500px")
```

Senate Districts with the largest total acreage of faith-owned parcels are Senate Districts 26 (12,598.90 acres) and 3 (6,381.74 acres). Once again, Senate District 26 is impacted by the presence of Shrine Mont.

```{r}
#| label: fig-acres-sd
#| fig-cap: "Total acres by Senate District"


# sd <- read_rds("data/rds/faith_senate.rds")
# 
# # Function to create color palette based on Jenks breaks
# create_jenks_palette <- function(data, n_classes = 5) {
#   # Remove any NA values
#   data <- data[!is.na(data)]
#   
#   # Calculate Jenks breaks
#   breaks <- classIntervals(data, n = n_classes, style = "jenks")
#   
#   # Create a color palette
#   pal <- colorBin(
#     palette = "YlGn",
#     domain = data,
#     bins = breaks$brks,
#     pretty = FALSE
#   )
#   
#   return(pal)
# }
# 
# # Create the Jenks breaks color palette
# jenks_pal <- create_jenks_palette(sd$faith_senate_total_acres)
# 
# # Create the leaflet map
# map_jenks <- leaflet(sd) %>%
#   addProviderTiles("CartoDB.Positron") %>%
#   addPolygons(
#     fillColor = ~jenks_pal(faith_senate_total_acres),
#     fillOpacity = 0.7,
#     color = "grey",
#     weight = 1,
#     popup = ~paste("Senate District:", NAMELSAD, "<br>Value:", label_acres)
#   ) %>%
#   addLegend(
#     pal = jenks_pal,
#     values = sd$faith_senate_total_acres,
#     title = "Total Acres",
#     position = "topleft"
#   )
# 
# # Display the map
# saveWidget(map_jenks, "data/geo/sd_acres.html")

knitr::include_url("data/geo/sd_acres.html", height = "500px")

```

Senate Districts with larger median parcel sizes are around the Richmond region and Hampton Roads region. This is a change from House Districts, where House District 28 was ranked fairly high. The increase in the area of the Senate District likely contributed to the change in median parcel size, with the inclusion of more smaller sized parcels. Senate Districts 12 (3.34 acres), 11 (2.88 acres), and 7 (2.05 acres) rank highest among their peers in terms of typical parcel size.

```{r}
#| label: fig-size-sd
#| fig-cap: "Median parcel size by Senate District"


# sd <- read_rds("data/rds/faith_senate.rds")
# 
# # Function to create color palette based on Jenks breaks
# create_jenks_palette <- function(data, n_classes = 5) {
#   # Remove any NA values
#   data <- data[!is.na(data)]
#   
#   # Calculate Jenks breaks
#   breaks <- classIntervals(data, n = n_classes, style = "jenks")
#   
#   # Create a color palette
#   pal <- colorBin(
#     palette = "YlGnBu",
#     domain = data,
#     bins = breaks$brks,
#     pretty = FALSE
#   )
#   
#   return(pal)
# }
# 
# # Create the Jenks breaks color palette
# jenks_pal <- create_jenks_palette(sd$med_parcel_size)
# 
# # Create the leaflet map
# map_jenks <- leaflet(sd) %>%
#   addProviderTiles("CartoDB.Positron") %>%
#   addPolygons(
#     fillColor = ~jenks_pal(med_parcel_size),
#     fillOpacity = 0.7,
#     color = "grey",
#     weight = 1,
#     popup = ~paste("Senate District:", NAMELSAD, "<br>Value:", label_size)
#   ) %>%
#   addLegend(
#     pal = jenks_pal,
#     values = sd$med_parcel_size,
#     title = "Median Parcel Size",
#     position = "topleft"
#   )
# 
# # Display the map
# saveWidget(map_jenks, "data/geo/sd_size.html")

knitr::include_url("data/geo/sd_size.html", height = "500px")

```

The data presented above represents a snapshot of faith-owned parcels by Virginia General Assembly Districts. A full exploration of these parcels for site feasibility is outside the scope of this report. However, it is clear that a significant amount of land exists that could represent potential residential development.

## Northern Virginia

The Virginia Zoning Atlas presents a unique opportunity to leverage research already conducted by HousingForward Virginia and the Mercatus Center at George Mason University. With the Virginia Zoning Atlas, parcels identified can be associated with their respective zoning districts and the rules governing residential development on them.

Conducting a spatial join in QGIS between the 652 identified parcels and the Northern Virginia Zoning Atlas resulted in 648 parcels retained. The four parcels that were removed were located in areas not covered by the Virginia Zoning Atlas, most likely to the area being deemed protected or undevelopable.

The graph below shows the number of faith-owned parcels in each jurisdiction within the Northern Virginia Regional Commission's footprint. Prince William County contains the most faith-owned parcels at 214, followed by Alexandria at 107. The fewest parcels are mainly within smaller towns, but also the City of Manassas Park.

```{r}
#| label: fig-nova-count
#| fig-cap: "Total parcels by jurisdiction in the NVRC region"

nova_zoning <- st_read("data/nova_faith_zoning.gpkg", quiet = TRUE)|> 
  st_drop_geometry() |> 
  distinct(ll_uuid, .keep_all = TRUE)


count_type_area <- nova_zoning |> 
  drop_na(jurisdiction_2) |> 
  group_by(jurisdiction_2) |> 
  summarise(acres = sum(acres),
            count = n_distinct(ll_uuid)
            )

ggplot(count_type_area,
       aes(x = reorder(jurisdiction_2, count),
           y = count)) +
  geom_col(
    fill = "#40c0c0"
  ) +
  geom_text(aes(
    label = comma(count), 
    hjust = -0.1
    )) +
  coord_flip() +
  theme_hfv(
    base_size = 10
    ) +
  scale_y_continuous(
    labels = number_format(big.mark = ","),
    expand = expansion(mult = c(0, 0.01)),
    limits = c(0, 250)) +
  labs(title = "Total faith-owned parcels by jurisdiction")

```

Prince William County maintains its top spot when looking at total acreage of those faith-owned parcels (1,213.50 acres). But this time, they are followed by Loudoun County, who comes in at 191.75 acres, only a sixth of the acreage in Prince William. All other jurisdictions in the Northern Virginia region have a total acreage under 100 for faith-owned parcels.

```{r}
#| label: fig-nova-acres
#| fig-cap: "Total acres by jurisdiction in the NVRC region"

nova_zoning <- st_read("data/nova_faith_zoning.gpkg", quiet = TRUE)|> 
  st_drop_geometry() |> 
  distinct(ll_uuid, .keep_all = TRUE)

count_type_area <- nova_zoning |> 
  drop_na(jurisdiction_2) |> 
  group_by(jurisdiction_2) |> 
  summarise(acres = round(sum(acres), digits = 2),
            count = n_distinct(ll_uuid)
            ) 

ggplot(count_type_area,
       aes(x = reorder(jurisdiction_2, acres),
           y = acres)) +
  geom_col(
    fill = "#40c0c0"
  ) +
  geom_text(aes(
    label = comma(acres), 
    hjust = -0.1
    )) +
  coord_flip() +
  theme_hfv(
    base_size = 10
    ) +
  scale_y_continuous(
    labels = number_format(big.mark = ","),
    expand = expansion(mult = c(0, 0.01)),
    limits = c(0, 1500)
    ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(linewidth = 0.5)
    ) +
  labs(
    title = "Total acres of faith-owned parcels by jurisdiction")
```

The Virginia Zoning Atlas allows users and researchers to understand what types of housing can be built, and whether what can be built requires a public hearing process. This process called "entitlement" is a process that requires property owners and developers to meet certain requirements and receive approval from the governing body to build what it is proposing. There are important considerations that go into these decisions, such as whether what is being proposed aligns with the community's comprehensive plan or whether the existing community opposes it or not. This process can be cumbersome and illicit irrational opposition based on misconceptions about density and change. Elected officials can easily be pushed by a vocal minority to vote down housing in spite of its alignment with community plans and explicit need.

The interactive chart below shows a breakdown of each jurisdiction in the Northern Virginia region in terms of:

- Total count of faith-owned parcels
- Percentage of those parcels that can be developed as single-family housing
- Percentage of those parcels that can be developed as two-family housing
- Percentage of those parcels that can be developed as three or more family housing 

```{r}
#| label: tbl-pct-1
#| tbl-cap: "Percent of parcels zoned by-right by residential use and jurisdiction"

library(DT)

percent_to_numeric <- function(x) {
  as.numeric(sub("%", "", x)) / 100
}

pct_tbl <- read_rds("data/rds/nova_byright_table.rds") |> 
  drop_na(jurisdiction) |> 
  mutate(across(3:5, .fns = ~ percent_to_numeric(.x)))

# pct_tbl |> 
#   arrange(desc(parcels)) |> 
#   kbl(col.names = c("Jurisdiction", "Parcel Count", "Single-Family", "Duplex", "Multifamily (3+)"),
#       align = "lcccc") |> 
#     kable_styling(bootstrap_options = c("striped", "condensed"))



datatable(pct_tbl, 
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; color: grey; font-size: 20px; 
            font-family: "Open Sans", sans-serif;
            padding-bottom: 20px;
            font-weight: bold;
            ',
            'Percent of parcels zoned by-right'
          ),

          colnames = c("Jurisdiction", "Parcel Count", "Single-Family",
                       "Duplex", "Multifamily"),
          rownames = FALSE,
          options = list(
            pageLength = -1,  # Show all rows
            dom = 't',        # Remove search bar and other controls
            scrollY = "400px",  # Add vertical scrolling if needed
            scrollCollapse = TRUE,
            initComplete = JS(
              "function(settings, json) {",
              "$(this.api().table().header()).css({'font-family': 'Open Sans'});",
              "}"
          ),
            columnDefs = list(list(className = 'dt-center', targets = 1:4))  # Center-align columns 2-5
          ),
          style = "bootstrap",
          fillContainer = TRUE) %>%
  formatStyle(
    columns = 3,  # Single-Family
    background = styleColorBar(pct_tbl[, 3], '#40c0c0'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center',
    color = '#606060'  # Set text color to light grey
  ) %>%
  formatStyle(
    columns = 4,  # Duplex
    background = styleColorBar(pct_tbl[, 4], '#8b85ca'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center',
    color = '#606060'  # Set text color to light grey
  ) %>%
  formatStyle(
    columns = 5,  # Multifamily
    background = styleColorBar(pct_tbl[, 5], '#011e41'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center',
    color = '#606060'  # Set text color to light grey
  ) %>%
  formatPercentage(columns = 3:5, digits = 1) |> 
    formatStyle(names(pct_tbl),
              fontFamily = "Open Sans")




```

For a majority of parcels in the region, many could be redeveloped as single-family housing without having to go through a public hearing process. However, fewer and fewer parcels can be redeveloped into denser housing options like duplexes or multifamily housing. Alexandria and Arlington are the most permissive due to their most recent zoning changes, but they have faced legal challenges that are put the analysis into question in those areas. More recently a judge overturned the Missing Middle zoning changes in Arlington and these have no doubt impacted the analysis. What will come of the legal challenges in Alexandria is yet to be seen, but the Arlington decision certainly puts the permissiveness on faith-owned parcels in Alexandria in jeopardy.

The figure below specifically focuses on by-right multifamily housing. As mentioned before, the number for Alexandria could change pending a court case, while the Arlington number will have certainly changed and will require a reversion to the previous zoning ordinance without the most recent Missing Middle amendments. Following these two localities, the City of Falls Church allows for 44 percent of identified faith-owned parcels to be developed as multifamily housing by-right. These figures decrease across the remaining jurisdictions with some not allowing for any faith-owned parceled to be developed by-right as multifamily housing.

```{r}
#| label: fig-pct-2
#| fig-cap: "Percent of parcels zoned by-right for multifamily housing"


pct_bar <- read_rds("data/rds/nova_byright_table.rds") |> 
  drop_na(jurisdiction) |> 
  mutate(multifamily = (parse_number(multifamily))/100)

ggplot(pct_bar,
       aes(x = reorder(jurisdiction, multifamily),
           y = multifamily)) +
  geom_col(
    fill = "#40c0c0"
  ) +
  geom_text(aes(
    label = percent(multifamily, 2)), 
    hjust = -0.1
    ) +
  coord_flip() +
  theme_hfv(
    base_size = 10
    ) +
  scale_y_continuous(
    labels = percent_format(),
    expand = expansion(mult = c(0, 0.01)),
    limits = c(0, 1.01)
    ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(linewidth = 0.5)
    ) +
  labs(title = "Percent of faith-owned parcels zoned for by-right multifamily")

```


## Conclusions

This report represents a preliminary examination of faith-owned parcels in the Commonwealth of Virginia. With additional time and resources, a team of researchers could conduct a deeper dive to determine the potential development capacity of parcels. Such an analysis could yield a more accurate sense of the impact of the Faith in Housing Bill should property owners choose to exercise a new right to develop. 

Nonetheless, the findings here represent the first attempt to identify parcels owned by faith-based organizations within the entire Commonwealth of Virginia. Over 20,000 parcels that amount to over 74,000 acres represents the potential to create two cities worth of housing should all parcels be viable for development. 
